---
title: "Single cell analysis (10X tumors)"
author: Ludwig Geistlinger, Sehyun Oh, and Levi Waldron
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 2
vignette: >
  % \VignetteIndexEntry{Single cell analysis}
  % \VignetteEngine{knitr::rmarkdown}
---

# Setup

Dependencies:
```{r setup, message = FALSE}
# OVC
library(subtypeHeterogeneity) 
library(consensusOV)

# Single cell
library(scater)
library(scran)
library(DropletUtils)
library(scRNAseq)
library(SingleR)
library(slingshot)
library(monocle3)

# Annotation
library(EnsDb.Hsapiens.v86)
library(org.Hs.eg.db)
library(EnrichmentBrowser)

# Plotting
library(ComplexHeatmap)
library(ggpubr)
```

Constants:
```{r}
SUBTYPES <- c("DIF", "IMR", "MES", "PRO")
CELL.TYPES <- c("EPI", "LYMPH", "MYE", "STROM", "ENDO")
```

Colors
```{r cols}
cb.pink <- "#CC79A7"
cb.red <- "#D55E00"
cb.blue <- "#0072B2"
cb.yellow <- "#F0E442"
cb.green <- "#009E73"
cb.lightblue <- "#56B4E9"
cb.orange <- "#E69F00"

stcols <- c(cb.lightblue, cb.green, cb.orange, cb.pink) 
names(stcols) <- SUBTYPES
```

Parallel computation:
```{r}
bp <- BiocParallel::registered()[[1]]
```

# Preprocessing

Read 10X output (required files: `matrix.mtx`, `barcodes.tsv`, `genes.tsv`)
```{r}
sce <- DropletUtils::read10xCounts("tumor90")
dim(sce)
```

Annotating the rows:
```{r}
rownames(sce) <- scater::uniquifyFeatureNames(rowData(sce)$ID,
                                              rowData(sce)$Symbol)
location <- AnnotationDbi::mapIds(EnsDb.Hsapiens.v86, keys = rowData(sce)$ID, 
                                  column = "SEQNAME", keytype = "GENEID")
```

Testing for deviations from ambient expression:
```{r}
bcrank <- DropletUtils::barcodeRanks(counts(sce))
uniq <- !duplicated(bcrank$rank)

plot(bcrank$rank[uniq], bcrank$total[uniq], log="xy",
    xlab="Rank", ylab="Total UMI count", cex.lab=1.2)

abline(h=bcrank$inflection, col="darkgreen", lty=2)
abline(h=bcrank$knee, col="dodgerblue", lty=2)

legend("bottomleft", legend=c("Inflection", "Knee"),
    col=c("darkgreen", "dodgerblue"), lty=2, cex=1.2)
```

Quality control on the cells:
```{r}
df <- scater::perCellQCMetrics(sce, 
                       subsets = list(Mito = which(location == "MT")),
                       BPPARAM = bp)
par(mfrow = c(1,3))
hist(df$sum/1e3, xlab="Library sizes (thousands)", main="",
    breaks=20, col="grey80", ylab="Number of cells")
hist(df$detected, xlab="Number of expressed genes", main="",
    breaks=20, col="grey80", ylab="Number of cells")
hist(df$subsets_Mito_percent, xlab="Mitochondrial proportion (%)",
    ylab="Number of cells", breaks=20, main="", col="grey80")
par(mfrow = c(1,1))

high.mito <- scater::isOutlier(df$subsets_Mito_percent, nmads = 3, type = "higher")
libsize.drop <- scater::isOutlier(df$sum, nmads = 1, type = "lower", log = TRUE)
feature.drop <- scater::isOutlier(df$detected, nmads = 1, type = "lower", log = TRUE)
sce <- sce[,!(high.mito | libsize.drop | feature.drop)]

df <- data.frame(ByHighMito = sum(high.mito),
           ByLibSize = sum(libsize.drop),
           ByFeature = sum(feature.drop),
           Remaining = ncol(sce))

df
dim(sce)
```

Examining gene expression:
```{r}
ave <- scater::calculateAverage(sce, BPPARAM = bp)
hist(log10(ave), breaks = 100, main = "", col = "grey",
    xlab = expression(Log[10]~"average count"))
```

Remove genes that have too low average counts
```{r}
rowData(sce)$AveCount <- ave
to.keep <- ave > 0.001
sce <- sce[to.keep,]

# Number excluded genes
sum(!to.keep)
dim(sce)
```

Library-size normalization:
```{r}
clusters <- scran::quickCluster(sce, method = "igraph", 
                                min.mean = 0.1, BPPARAM = bp)
sce <- scran::computeSumFactors(sce, min.mean = 0.1,
                                cluster = clusters, BPPARAM = bp)
plot(scater::librarySizeFactors(sce), sizeFactors(sce), pch = 16,
    xlab = "Library size factors", ylab = "Deconvolution factors", log = "xy")
sce <- scater::logNormCounts(sce)
sce
```

# Dimension reduction

Modelling the mean-variance trend:

```{r}
dec.pbmc <- scran::modelGeneVarByPoisson(sce, BPPARAM = bp)
top.pbmc <- scran::getTopHVGs(dec.pbmc, prop = 0.1)
plot(dec.pbmc$mean, dec.pbmc$total, pch=16, cex=0.5,
    xlab="Mean of log-expression", ylab="Variance of log-expression")
curfit <- metadata(dec.pbmc)
curve(curfit$trend(x), col = 'dodgerblue', add = TRUE, lwd = 2)
```

Denoise log-expression data by removing principal components corresponding to 
technical noise (might take a while):

```{r}
sce <- scran::denoisePCA(sce, subset.row = top.pbmc,
                         technical = dec.pbmc, BPPARAM = bp)
ncol(reducedDim(sce, "PCA"))
plot(attr(reducedDim(sce), "percentVar"), xlab = "PC",
     ylab = "Proportion of variance explained")
abline(v = ncol(reducedDim(sce, "PCA")), lty = 2, col = "red")
```

tSNE:
```{r}
sce <- scater::runTSNE(sce, dimred = "PCA", perplexity = 30, BPPARAM = bp)
sce$sizeFactor <- sizeFactors(sce)
scater::plotTSNE(sce, colour_by = "sizeFactor")
```

# Clustering

Clustering with graph-based methods:
```{r}
snn.gr <- scran::buildSNNGraph(sce, use.dimred = "PCA", BPPARAM = bp, k = 25)
clusters <- igraph::cluster_walktrap(snn.gr)
sce$Cluster <- factor(clusters$membership)
table(sce$Cluster)
```

```{r}
cluster.mod <- scran::clusterModularity(snn.gr, sce$Cluster,
                                        get.weights = TRUE)
log.ratio <- log2(cluster.mod$observed / cluster.mod$expected + 1)

ComplexHeatmap::Heatmap(log.ratio,
                        cluster_rows = FALSE, cluster_columns = FALSE)
scater::plotTSNE(sce, colour_by = "Cluster")
```

# Annotation (cell type and consensus subtype) 

## SingleR cell type annotation

Transfer cell type annotation from reference profiles in 
the Human Primary Cell Atlas and ENCODE:
```{r, message = FALSE}
hpc <- transferCellType(sce, "hpca")
encode <- transferCellType(sce, "encode")

sce$hpca.celltype <- hpc$labels 
sce$encode.celltype <- encode$labels

maxScore <- function(res)
    vapply(1:nrow(res), function(i) res$scores[i, res[i,"labels"]], numeric(1))

sce$hpca.celltype.score <- maxScore(hpc)
sce$encode.celltype.score <- maxScore(hpc)
```

Display main cell types present:
```{r}
plotMainCellTypes(sce, "hpca.celltype")
plotMainCellTypes(sce, "encode.celltype")
```

Display ambiguity of cell type assignments:
```{r}
SingleR::plotScoreHeatmap(hpc)
SingleR::plotScoreHeatmap(encode)
```

## Marker genes

Adopted from Shih et al., PLoS One (2018):
```{r}
epithelial <- c("EPCAM", "KRT8", "KRT18", "KRT19")
lymphocyte <- c("PTPRC", "CD3E", "CD19", "MS4A1")
endothelial <- c("PECAM1", "CD34")
fibroblast <- c("ACTA2", "DCN", "ACTB")
stromal <- c("THY1", "ENG", "VIM", "CD44")

origin <- c("PAX8", "CALB2", "COL11A1")
scater::plotTSNE(sce, colour_by = epithelial[1])
```

## consensusOV subtype annotation

```{r}
sce.entrez <- EnrichmentBrowser::idMap(sce, org = "hsa", 
                                       from = "SYMBOL", to = "ENTREZID")
am <- as.matrix(assay(sce.entrez, "logcounts"))
cst <- consensusOV::get.consensus.subtypes(am, names(sce.entrez))

sts <- sub("_consensus$", "", as.vector(cst$consensusOV.subtypes))
sce$subtype <- sts
sce$margin <- consensusOV::margin(cst$rf.probs)

summary(sce$margin)
scater::plotTSNE(sce, colour_by = "subtype")
scater::plotTSNE(sce, colour_by = "margin")
```

## Visualization of results (all tumors)

Read the pre-processed 10X data for all tumors as `list` of `SingleCellExperiment`s. 
```{r}
tumors <- c(59, 76, 77, 89, 90) 
readSCE <- function(i) readRDS(gsub("XX", i, "tumorXX/sampleXX_sce.rds")) 
sces <- lapply(tumors, readSCE)
names(sces) <- paste0("T", tumors)
sces
```

Summarize cell type annotation:
```{r, eval = FALSE}
sces <- lapply(sces, annoCellType)
```

```{r}
vapply(sces, function(sce) table(sce$celltype), integer(5))
vapply(sces, function(sce) table(sce$subtype), integer(4))
```

Display cell type and subtype annotation for one tumor at a time:
```{r, fig.height = 6, fig.wide = TRUE}
p2a <- plotType(sces[[1]])
p2b <- plotType(sces[[1]], type = "subtype")
p2 <- ggarrange(p2b + theme_bw() + theme(plot.margin = margin(r = 1)), 
                p2a + theme_bw() + theme(axis.text.y = element_blank(), 
                                         axis.ticks.y = element_blank(), 
                                         axis.title.y = element_blank(),
                                         plot.margin = margin(l = 1, r=1)), 
                legend = "top", align = "h", labels = c("A", "B"))
p2
```
or
```{r, fig.height = 6, fig.small = TRUE}
p3 <- ggarrange(p2b + theme_bw() + 
                      theme(axis.text.x = element_blank(),  
                            axis.ticks.x = element_blank(), 
                            axis.title.x = element_blank()),  
                p2a + theme_bw(),   
                nrow = 2, legend = "none", align = "hv", labels = c("A", "B"))
p3
```

Facet cell type and subtype visualization for all tumors:
```{r, fig.wide = TRUE, fig.height = 6}
stp <- facetTumors(sces, col = "subtype")
pal <- get_palette("npg", 5)
names(pal) <- CELL.TYPES
ctp <- facetTumors(sces, col = "celltype", pal = pal)
ggarrange(stp, ctp, nrow = 2, align = "hv")
```

Barplot percentage of subtype / cell type for all 10X tumors:

```{r}
cttab <- vapply(sces, getPerc, numeric(5), perc=FALSE)
total <- colSums(cttab)
prop.test(cttab["EPI",], total)
prop.test(cttab["STROM",], total)
```

```{r, fig.height = 6}
stbp <- bpType(sces, "subtype")
ctbp <- bpType(sces)
p1 <- ggarrange(stbp + theme(axis.text.x = element_blank(),
                             axis.title.x = element_blank()),
                             # axis.ticks.x = element_blank(),
                             # axis.line.x = element_blank()), 
                ctbp, ncol = 1,
                legend = "left", align = "hv", labels = c("C", "D"))
p1
```

Cell type vs subtype matrix:
```{r}
mat <- getCelltypeSubtypeMatrix(sces)
ctstbp <- bpCrossType(mat)
ctstbp
```

Or as matrix plot:
```{r}
mat <- round(mat / sum(mat) * 100)
mat <- mat[,CELL.TYPES]
matrixPlot(mat, cnames = c("SUBTYPE", "CELL.TYPE"), high.color = "darkgreen")
```

Margin scores:

```{r}
spl.margin <- function(sce, col = "celltype")
    reshape2::melt(split(sce$margin, sce[[col]]))
psces <- lapply(sces, spl.margin)
for(i in tumors) psces[[match(i, tumors)]]$tumor <- paste0("T", i)
df <- do.call(rbind, psces)
colnames(df) <- c("MARGIN", "CELL.TYPE", "TUMOR")
df[[2]] <- factor(df[[2]], levels = CELL.TYPES)
```

Boxplot of margin scores:
```{r, fig.height = 4}
ggboxplot(df, x = "CELL.TYPE", y = "MARGIN", width = 0.8, notch = TRUE,
            fill = "CELL.TYPE", palette = get_palette("lancet", 5),
            facet.by = "TUMOR", nrow = 1,  
            x.text.angle = 45, legend="none", ggtheme = theme_bw(base_size = 12)) 
```

Facet TSNEs of margin scores:
```{r, fig.height = 4}
facetTumors(sces, "margin", cont = TRUE)
```

Only epithelial cells:
```{r}
df <- subset(df, CELL.TYPE == "EPI")
df <- subset(df, TUMOR != "T61")
bp <- ggboxplot(df, x = "TUMOR", y = "MARGIN", width = 0.8, notch = TRUE, 
                fill = cb.red, 
                legend="none", ggtheme = theme_bw(), ylab = "MARGIN (EPI CELLS)",
                font.x = 14, font.y = 14, font.tickslab = c(13, "plain", "black"))
                # ylim = c(0,1),x.text.angle = 45, 
df.points <- data.frame(x=1:4, y=c(0.376, 0.766, 0.182, 0.646))
bp <- bp + geom_point(data = df.points, aes(x = x, y = y), shape = 8, size = 3) +
           geom_text(data = df.points[2,], aes(x = x, y = y),
                     label = "Bulk", hjust = 0, nudge_x = 0.1, size = 5)
bp
```

## Bulk RNA-seq

```{r}
dat1 <- read.delim("Bulk/Bulk_RNAseq.txt", as.is=TRUE)
dat <- read.delim("Bulk/Bulk_RNAseq2.txt", as.is=TRUE)
cpm.ind <- seq_len(which(dat[,1] == "")[1] - 1)
cpm.dat <- dat[cpm.ind, 1:8]
raw.dat <- dat[,9:16]
cpm.sym <- cpm.dat[,1]
raw.sym <- raw.dat[,1]
```

Use pre-computed library size-normalized expression values for subtype assignment:
```{r, message = FALSE}
rownames(cpm.dat) <- cpm.dat[,2] 
cpm.dat <- cpm.dat[,3:8]
colnames(cpm.dat) <- paste0("T", c(59, 77, 76, 89, 90, 91)) 
cpm.dat <- as.matrix(cpm.dat)
mode(cpm.dat) <- "numeric"
cpm.se <- SummarizedExperiment(assays = list(cpm = cpm.dat))
cpm.se <- EnrichmentBrowser::idMap(cpm.se, org="hsa", from="ENSEMBL", to="ENTREZID")
# cst <- consensusOV::get.consensus.subtypes(assay(cpm.se), names(cpm.se))
```

```{r}
cst <- readRDS("Bulk/cpm_subtypes.rds")
cst
```

Sanity check: start from raw read counts
```{r, message = FALSE}
rownames(raw.dat) <- raw.dat[,2]
raw.dat <- raw.dat[,3:8]
colnames(raw.dat) <- paste0("T", c(59, 77, 76, 89, 90, 91))
raw.dat <- as.matrix(raw.dat)
mode(raw.dat) <- "numeric"
raw.dat <- edgeR::cpm(raw.dat, log=TRUE)
raw.se <- SummarizedExperiment(assays = list(raw = raw.dat))
raw.se <- EnrichmentBrowser::idMap(raw.se, org="hsa", from="ENSEMBL", to="ENTREZID")
# cst2 <- consensusOV::get.consensus.subtypes(assay(raw.se), names(raw.se))
```

Pretty much the same as when starting from already normalized expression values:
```{r}
cst2 <- readRDS("Bulk/raw_subtypes.rds")
cst2
```

Contrast bulk and single-cell margins:
```{r, fig.height = 8}
mat <- t(cst$rf.probs)
rownames(mat) <- sub("_consensus", "", rownames(mat))
mat <- round(mat, digits = 2)
mat <- mat[c(3,4,1,2),c(1,3,2,4)]
mp <- matrixPlot(t(mat), cnames = c("TUMORS", "CLASS.PROB"), 
                    bw.thresh = 0.5, gthresh = 0.2)
mp <- mp + theme(axis.text.x = element_blank(),
                 axis.title.x = element_blank(),
                 axis.ticks.x = element_blank(),
                 axis.line.x = element_blank(),
                 plot.margin = margin(b = 0))
mp <- mp + ylab("BULK")
p0 <- ggarrange(ctstbp, mp, bp, 
                nrow = 3, align = "v", heights = c(2,1,3), labels = c("E", "F"))
p0
```

Heatmaps for transitional DIF-PRO patterns on EPI cells (Tumors 59 and 77):
```{r, eval = FALSE}
sces <- lapply(tumors, getClassProbs)
```

```{r, fig.height = 9, fig.wide = TRUE}
h1 <- scHeatmap2(sces[[1]], title = "Cells (T59)", fsize = 8)
h2 <- scHeatmap2(sces[[4]], title = "Cells (T77)", fsize = 8, 
                name = "h2", add = TRUE, legend = FALSE)
h1 + h2
for(st in names(stcols)) 
    ComplexHeatmap::decorate_annotation(st,
        {grid::grid.text(st, grid::unit(-2, "mm"), just = "right")})
```

# InferCNV

- tumor 59: 3,500 epithelial cells vs. 9,500 reference (non-epithelial) cells 
- tumor 76: 210 epithelial cells vs. 13,500 reference (non-epithelial) cells 
- tumor 77: 2,900 epithelial cells vs. 4,000 reference (non-epithelial) cells 
- tumor 89: 300 epithelial cells vs. 4,600 reference (non-epithelial) cells 
- tumor 90: 900 epithelial cells vs. 2,700 reference (non-epithelial) cells 

## Prepare input files and execute InferCNV

Prepare input files from a `SingleCellExperiment`
```{r, eval = FALSE}
obj <- generateInferCNVInput(sces[[1]], "tumor59")
```

Run `inferCNV` for one tumor at a time (computational intensive, several hours).
Best executed on a cluster (here with 20 cores).

```{r, eval = FALSE}
# cutoff = 1 works well for Smart-seq2, 
# cutoff = 0.1 works well for 10x Genomics
res <- infercnv::run(obj,
                     cutoff = 0.1, 
                     out_dir = "tumor59",
                     num_threads = 20,
                     denoise = TRUE,
                     HMM = TRUE)
```

## Cross-tumor heatmap

```{r}
for(i in seq_along(sces)) colnames(sces[[i]]) <- sces[[i]]$Barcode
```

Read observation matrix (expression profiles of cancer epithelial cells)
```{r, message = FALSE}
obs.mats <- lapply(tumors, readObservations)
```

Subset and annotate cells:
```{r}
for(i in seq_along(tumors)) 
    obs.mats[[i]] <- subsetObservations(obs.mats[[i]], sces[[i]])
ra <- getCellAnnotation(sces, obs.mats)
```

Create union of genes featured in the inferCNV output of each tumor, 
extend individual observations as needed, and create a combined observation matrix
storing the information across all tumors:
```{r}
un <- Reduce(union, lapply(obs.mats, rownames))
obs.mats <- lapply(obs.mats, extendObservations, un = un) 
obs.mat <- do.call(cbind, lapply(obs.mats, function(o) o[un,]))
```

Obtain genomic coordinates for the genes in the combined observation matrix:
```{r, message = FALSE}
egenes <- symbols2ranges(rownames(obs.mat))
obs.mat <- obs.mat[names(egenes),]
chr <- as.character(seqnames(egenes))
```

Highlight specific recurrent regions with known driver genes from the TCGA OVC paper:
```{r, message = FALSE}
clabs <- getRecurrentDriverGenes(rownames(obs.mat))

```

Plot the heatmap:
```{r, fig.wide = TRUE, fig.height = 10}
ComplexHeatmap::Heatmap(t(obs.mat),
    #top_annotation = ta,
    right_annotation = ra,
    name = "Expression",
    col = circlize::colorRamp2(c(0.895, 1, 1.105), c("blue", "white", "red")),
    row_title_rot = 0, column_title_rot = 90,
    show_row_names = FALSE, show_column_names = TRUE,
    cluster_rows = FALSE, cluster_columns = FALSE,
    row_split = rep(paste0("T", tumors), sapply(obs.mats, ncol)),
    column_spl = factor(chr, levels = unique(chr)),
    column_gap = unit(0, "mm"),
    column_title_gp = gpar(fontsize = 6),
    column_labels = clabs,
    column_names_gp = gpar(fontsize = 6),
    border = TRUE)
```

## Tumor 77

Tumor 77 shows an unusual bimodal distribution with regard to number of genes
expressed per cell:

```{r}
sce77 <- sces$T77
ggpubr::gghistogram(sce77$total_features_by_counts, bins = 25, 
            xlab = "number of genes expressed", 
            ylab = "number of cells",
            fill = "#00AFBB", ggtheme = ggplot2::theme_bw())
```

There is clear effect:
```{r}
om77 <- obs.mats[[3]]
colnames(sce77) <- colData(sce77)$Barcode
nrg <- sce77[,colnames(om77)]$total_features_by_counts
ca <- ComplexHeatmap::rowAnnotation(nr.genes = nrg)
ComplexHeatmap::Heatmap(t(om77), 
    col = circlize::colorRamp2(c(0.895, 1, 1.105), c("blue", "white", "red")), 
    name = "Expression", right_annotation = ca, 
    cluster_rows = FALSE, cluster_columns = FALSE, 
    show_row_names = FALSE, show_column_names = FALSE, 
    column_title = "Genes", row_title = "Cells")
```

Split into two partitions:
```{r}
nrg <- sce77$total_features_by_counts
sce77.high <- sce77[,nrg > 2000]
sce77.low <- sce77[,nrg < 2000]
```

For the partition with low overall expression, be more restrictive on cell type calling:
```{r}
ctype <- sce77.low$celltype
lcounts <- assay(sce77.low, "logcounts")
sce77.low$celltype[ctype == "EPI" & lcounts["EPCAM",] < 1] <- NA  
sce77.low$celltype[ctype == "LYMPH" & lcounts["PTPRC",] < 1] <- NA
sce77.low$celltype[ctype == "MYE" & lcounts["AIF1",] < 1] <- NA
sce77.low$celltype[ctype == "STROM" & lcounts["DCN",] < 1] <- NA
sce77.low$celltype[ctype == "ENDO" & lcounts["PECAM1",] < 1] <- NA
ctype <- sce77.low$celltype
```

Combined view on infercnv results for both partitions:
```{r, fig.height = 10}
obs77l <- readObservations("77l")
obs77h <- readObservations("77h")
obs.mats77 <- list(obs77h, obs77l)
sces77 <- list(sce77.high, sce77.low[,!is.na(ctype)])
colnames(sces77[[1]]) <- sces77[[1]]$Barcode
for(i in 1:2) obs.mats77[[i]] <- subsetObservations(obs.mats77[[i]], sces77[[i]])
ra <- getCellAnnotation(sces77, obs.mats77)
un <- Reduce(union, lapply(obs.mats77, rownames))
obs.mats77 <- lapply(obs.mats77, extendObservations, un = un)
obs.mat <- do.call(cbind, lapply(obs.mats77, function(o) o[un,]))
egenes <- symbols2ranges(rownames(obs.mat))
obs.mat <- obs.mat[names(egenes),]
chr <- as.character(seqnames(egenes))
clabs <- getRecurrentDriverGenes(rownames(obs.mat))
ComplexHeatmap::Heatmap(t(obs.mat),
     #top_annotation = ta,
     right_annotation = ra,
     name = "Expression",
     col = circlize::colorRamp2(c(0.895, 1, 1.105), c("blue", "white", "red")),
     row_title_rot = 0, column_title_rot = 90,
     show_row_names = FALSE, show_column_names = TRUE,
     cluster_rows = FALSE, cluster_columns = FALSE,
     row_split = rep(c("T77h", "T77l"), sapply(obs.mats77, ncol)),
     column_spl = factor(chr, levels = unique(chr)),
     column_gap = unit(0, "mm"),
     column_title_gp = gpar(fontsize = 6),
     column_labels = clabs,
     column_names_gp = gpar(fontsize = 6),
     border = TRUE)
``` 

Assess whether the low-expression partition is just an under-sampled part of the
high-expression partition, by (i) down-sampling EPI cells of the high-expression
partition to match the coverage of the EPI cells in the low-expression partition,

```{r}
am <- as.matrix(assay(sces77[[1]]))
ctype <- sces77[[1]]$celltype == "EPI"
high.ls <- summary(colSums(am[,ctype], na.rm = TRUE))
am2 <- as.matrix(assay(sces77[[2]]))
ctype2 <- sces77[[2]]$celltype == "EPI"
low.ls <- summary(colSums(am2[,ctype2], na.rm = TRUE))
(ls.prob <- low.ls["Median"] / high.ls["Median"])
high.down <- DropletUtils::downsampleMatrix(am[,ctype], prop=ls.prob)
```

and (ii) computing pairwise correlation for random pairs of cells of the 
low-expression partition and cells of the down-sampled high expression partition.

```{r, warning = FALSE}
am <- am2[,ctype2]
grid <- sample(ncol(high.down), ncol(am))
cvec <- vector("numeric", length = ncol(am))
for(i in seq_along(cvec)) cvec[i] <- cor(am[,i], high.down[,grid[i]])
summary(cvec)
```

```{r}
sces$T77 <- sce77.high  
```

# Cell Cycle

## Cyclins

### Tumor 59

```{r}
for(i in seq_along(sces)) colnames(sces[[i]]) <- NULL
sce59 <- subsetByCellType(sces[[1]], cell.type = "EPI", subtype = c("DIF", "PRO"))
sce59
```

```{r, fig.height = 6}
cyclin.genes <- grep("^CCN[ABDE][0-9]$", names(sce59), value = TRUE)
cyclin.genes
markerHeatmap(sce59, cyclin.genes)
bpCyclins(sce59)
```

### Tumor 77
```{r}
sce77 <- subsetByCellType(sces[[3]], cell.type = "EPI", subtype = c("DIF", "PRO"))
bpCyclins(sce77)
```

### Tumor 90
```{r}
sce90 <- subsetByCellType(sces[[5]], cell.type = "EPI", subtype = c("DIF", "PRO"))
bpCyclins(sce90)
```

## Marker genes

### Tumor 59

```{r, fig.height = 6}
markers59 <- scran::findMarkers(sce59, groups = sce59$subtype)
head(markers59$DIF)
markers59$DIF["CCND1",]
head(markers59$DIF[markers59$DIF$logFC.PRO < 0,])
markerHeatmap(sce59, rownames(markers59$DIF)[1:25])
```

Contrast with bulk markers
```{r, fig.height = 6}
data.dir <- system.file("extdata", package = "subtypeHeterogeneity")
verhaak.file <- file.path(data.dir, "Verhaak_supplementT8A.txt")
bulk.markers <- getExtendedVerhaakSignature(verhaak.file, 
                                            nr.genes.per.subtype = 200,
                                            by.subtype = TRUE)
bmarkers59 <- compileMarkers(markers59, bulk.markers)
bmarkers59
markerHeatmap(sce59, rownames(bmarkers59), row.split = TRUE)
```

### Tumor 77
```{r, fig.height = 6}
markers77 <- scran::findMarkers(sce77, groups = sce77$subtype)
markerHeatmap(sce77, rownames(markers77$DIF)[1:25])
bmarkers77 <- compileMarkers(markers77, bulk.markers)
markerHeatmap(sce77, rownames(bmarkers77), row.split = TRUE)
```

### Tumor 90
```{r, fig.height = 6}
markers90 <- scran::findMarkers(sce90, groups = sce90$subtype)
markerHeatmap(sce90, rownames(markers90$DIF)[1:25])
bmarkers90 <- compileMarkers(markers90, bulk.markers)
markerHeatmap(sce90, rownames(bmarkers90), row.split = TRUE)
```

### Combined

```{r}
markers <- list(markers59, markers77, markers90)
dif.markers <- getCommonMarkers(markers, n = 10)
pro.markers <- getCommonMarkers(markers, subtype = "PRO", n = 10)
markers <- c(dif.markers, pro.markers)
```

# capture the heatmap
```{r}
ml <- markerHeatmapList(sces[c(1,3,5)], markers, row.split = TRUE)
m <- grid.grabExpr(draw(ml))
```

## Annotation transfer from reference profiles
```{r, message = FALSE}
sce.ref <- scRNAseq::LengESCData()
sce.ref
```

## Cyclone 

### Tumor 59

Run the cyclone classifier (pre-computed, computationally intensive)
```{r, eval = FALSE}
sce.ensembl <- EnrichmentBrowser::idMap(sce, org = "hsa", 
                                        from = "SYMBOL", to = "ENSEMBL")
marker.file <- system.file("extdata", "human_cycle_markers.rds", package="scran")
hm.pairs <- readRDS(marker.file)
cy.assignments <- scran::cyclone(sce.ensembl, hm.pairs)
saveRDS(cy.assignments, file = "tumor59/cyclone.rds")
```

Get the computed assignments and plot:
```{r}
cy.assignments <- readRDS("tumor59/cyclone.rds")
plot(cy.assignments$score$G1, cy.assignments$score$G2M,
     xlab = "G1 score", ylab = "G2/M score", pch = 16)
df <- data.frame(cy.assignments$score, subtype = sce59$subtype)
ggscatter(df, x = "G1", y = "G2M", color = "subtype",
          palette = stcols[c("DIF", "PRO")], ggtheme = ggplot2::theme_bw())
```

```{r}
tab <- table(cy.assignments$phases, sce59$subtype)
tab
chisq.test(tab)
```

Figure 5:

```{r}
bc <- bpCyclins(sces[[1]], top.only = TRUE) + ylim(c(0,4.3))
bc <- geom_signif(annotations = "1.8e-07", 
                  y_position = 4.25, xmin = 4.8, xmax = 5.2) 
gs <- ggscatter(df, x = "G1", y = "G2M", 
                color = "subtype", palette = stcols[c("DIF", "PRO")], 
                ggtheme = ggplot2::theme_bw(), legend = "none")

# ggarrange(im, ggarrange(ggarrange(bc, gs, ncol = 2, labels = c("B", "C")), m, nrow = 2, labels = c("", "D")), ncol = 2, labels = c("A",""))
```

### Tumor 77

```{r}
cy.assignments <- readRDS("tumor77/cyclone.rds")
df <- data.frame(cy.assignments$score, subtype = sce77$subtype)
ggscatter(df, x = "G1", y = "G2M", color = "subtype",
          palette = stcols[c("DIF", "PRO")], ggtheme = ggplot2::theme_bw())
```

```{r}
tab <- table(cy.assignments$phases, sce77$subtype)
tab
chisq.test(tab)
```


### Tumor 90

```{r}
cy.assignments <- readRDS("tumor90/cyclone.rds")
df <- data.frame(cy.assignments$score, subtype = sce90$subtype)
ggscatter(df, x = "G1", y = "G2M", color = "subtype",
          palette = stcols[c("DIF", "PRO")], ggtheme = ggplot2::theme_bw())
```

```{r}
tab <- table(cy.assignments$phases, sce90$subtype)
tab
fisher.test(tab)
```

# Trajectory analysis

## Slingshot

### Tumor 77

Preprocess:
```{r}
ind <- which(tumors == 77)
sce77 <- subsetByCellType(sces[[ind]], cell.type = "EPI", subtype = c("DIF", "PRO"))
colData(sce77) <- colData(sce77)[,!grepl("slingPseudotime", colnames(colData(sce77)))]
sce77 <- preprocessSCE(sce77)

ncol(reducedDim(sce77, "PCA"))
table(sce77$Cluster)
scater::plotTSNE(sce77, colour_by = "Cluster")
scater::plotTSNE(sce77, colour_by = "subtype")
scater::plotTSNE(sce77, colour_by = "margin")
```

Run slingshot + inspect output
```{r}
#reducedDims(sce77)$PCA <- reducedDims(sce77)$PCA[,1:10]
sce77 <- slingshot::slingshot(sce77, clusterLabels = "Cluster",
                            reducedDim = "PCA", approx_points = 200)
grep("slingPseudotime", colnames(colData(sce77)), value = TRUE)

scater::plotTSNE(sce77, colour_by = "slingPseudotime_1")
scater::plotTSNE(sce77, colour_by = "slingPseudotime_2")
```

Which genes change significantly as a function of pseudotime?
```{r}
pseudotimeHeatmap(sce77, lineage = 1, nr.genes = 50)
```

### Tumor 59

Preprocess:
```{r}
ind <- which(tumors == 59)
sce59 <- subsetByCellType(sces[[ind]], cell.type = "EPI", subtype = c("DIF", "PRO"))
colData(sce59) <- colData(sce59)[,!grepl("slingPseudotime", colnames(colData(sce59)))]
sce59 <- preprocessSCE(sce59)

ncol(reducedDim(sce59, "PCA"))
table(sce59$Cluster)
scater::plotTSNE(sce59, colour_by = "Cluster")
scater::plotTSNE(sce59, colour_by = "subtype")
scater::plotTSNE(sce59, colour_by = "margin")
```

Run slingshot + inspect output
```{r}
#reducedDims(sce59)$PCA <- reducedDims(sce59)$PCA[,1:10]
sce59 <- slingshot::slingshot(sce59, clusterLabels = "Cluster",
                            reducedDim = "PCA", approx_points = 200,
                            start.clus = 4, end.clus = 5)
grep("slingPseudotime", colnames(colData(sce59)), value = TRUE)

scater::plotTSNE(sce59, colour_by = "slingPseudotime_1")
scater::plotTSNE(sce59, colour_by = "slingPseudotime_2")
```

Which genes change significantly as a function of pseudotime?
```{r}
pseudotimeHeatmap(sce59, lineage = 1, nr.genes = 50)
```

### Tumor 90

Preprocess:
```{r}
ind <- which(tumors == 90)
sce90 <- subsetByCellType(sces[[ind]], cell.type = "EPI", subtype = c("DIF", "PRO"))
colData(sce90) <- colData(sce90)[,!grepl("slingPseudotime", colnames(colData(sce90)))]
sce90 <- preprocessSCE(sce90)

ncol(reducedDim(sce90, "PCA"))
table(sce90$Cluster)
scater::plotTSNE(sce90, colour_by = "Cluster")
scater::plotTSNE(sce90, colour_by = "subtype")
scater::plotTSNE(sce90, colour_by = "margin")
```

Run slingshot + inspect output
```{r}
#reducedDims(sce90)$PCA <- reducedDims(sce90)$PCA[,1:10]
sce90 <- slingshot::slingshot(sce90, clusterLabels = "Cluster",
                            reducedDim = "PCA", approx_points = 200,
                            start.clus = 2, end.clus = 1)
grep("slingPseudotime", colnames(colData(sce90)), value = TRUE)

scater::plotTSNE(sce90, colour_by = "slingPseudotime_1")
```

Which genes change significantly as a function of pseudotime?
```{r}
pseudotimeHeatmap(sce90, lineage = 1, nr.genes = 50)
```

MALAT1 promotes proliferation and metastasis in epithelial ovarian cancer
via the PI3K-AKT pathway, Eur Rev Med Pharmacol Sci. 2017

NEAT1 promotes cell proliferation and migration 
in high-grade serous ovarian cancer, Cell Death & Disease, 2018

Mitochondria as biosynthetic factories for cancer proliferation, 
Cancer Metabolism, 2015

## Monocle3

### Tumor 77

Creating a `cell_data_set` (aka `Monocle3`'s SingleCellExperiment)

```{r}
ind <- which(tumors == 77)
sce77 <- subsetByCellType(sces[[ind]], cell.type = "EPI", subtype = c("DIF", "PRO"))

colnames(sce77) <- sce77$Barcode
rowData(sce77)$gene_short_name <- rownames(sce77)
cds <- monocle3::new_cell_data_set(as.matrix(assays(sce77)$counts),
     cell_metadata = as.data.frame(colData(sce77)),
     gene_metadata = as.data.frame(rowData(sce77)))
```

Preprocessing + Dimensionality reduction:
```{r, message = FALSE}
cds <- monocle3::preprocess_cds(cds)
cds <- monocle3::reduce_dimension(cds)
monocle3::plot_pc_variance_explained(cds)
monocle3::plot_cells(cds, 
                     label_groups_by_cluster = FALSE,
                     color_cells_by = "subtype")
monocle3::plot_cells(cds,
                     label_groups_by_cluster = FALSE,
                     color_cells_by = "margin")
```

Clustering + learning trajectory graph
```{r, message = FALSE}
cds <- monocle3::cluster_cells(cds)
cds <- monocle3::learn_graph(cds)
monocle3::plot_cells(cds,
                     color_cells_by = "subtype",
                     label_groups_by_cluster = FALSE,
                     label_leaves = FALSE)
```

Identify root cell as the one with lowest MARCKSL1 expression 
```{r, message = FALSE}
root.cell <- names(which.min(assay(sce77, "logcounts")["MARCKSL1",]))
cds <- monocle3::order_cells(cds, root_cells = root.cell)
monocle3::plot_cells(cds,
                     color_cells_by = "pseudotime",
                     label_groups_by_cluster = FALSE,
                     label_leaves = FALSE)
pt <- monocle3::pseudotime(cds)
```

Finding genes that change as a function of pseudotime:
```{r, eval = FALSE}
res <- monocle3::graph_test(cds, neighbor_graph="principal_graph", cores = 2)
res <- res[order(res$morans_test_statistic, decreasing = TRUE),]
head(res, n = 20)
monocle3::plot_cells(cds, genes = "MALAT1")
monocle3::plot_cells(cds, genes = "NEAT1")
```

Plot gene expression dynamics:
```{r, eval = FALSE}
lineage_cds <- cds[rownames(cds) %in% c("SLPI", "HMGA1"),]
monocle3::plot_genes_in_pseudotime(lineage_cds,
                         color_cells_by="subtype",
                         min_expr=0.5)
```

Force a trajectory through both partitions:

```{r}
cds <- monocle3::learn_graph(cds, use_partition = FALSE)
monocle3::plot_cells(cds,
                     color_cells_by = "subtype",
                     label_groups_by_cluster = FALSE,
                     label_leaves = FALSE)
```

### Tumor 59

```{r}
ind <- which(tumors == 59)
sce59 <- subsetByCellType(sces[[ind]], cell.type = "EPI", subtype = c("DIF", "PRO"))

colnames(sce59) <- sce59$Barcode
rowData(sce59)$gene_short_name <- rownames(sce59)
cds <- monocle3::new_cell_data_set(as.matrix(assays(sce59)$counts),
     cell_metadata = as.data.frame(colData(sce59)),
     gene_metadata = as.data.frame(rowData(sce59)))
```

```{r}
cds <- monocle3::preprocess_cds(cds)
cds <- monocle3::reduce_dimension(cds)
cds <- monocle3::cluster_cells(cds)
cds <- monocle3::learn_graph(cds, verbose = FALSE)
monocle3::plot_cells(cds,
                     color_cells_by = "subtype",
                     label_groups_by_cluster = FALSE,
                     label_leaves = FALSE)
```

Identify root cell as the one with lowest NEAT1 expression 

```{r, message = FALSE}
monocle3::plot_cells(cds, genes = "NEAT1")
root.cell <- names(which.min(assay(sce59, "logcounts")["NEAT1",]))
cds <- monocle3::order_cells(cds, root_cells = root.cell)
monocle3::plot_cells(cds,
                     color_cells_by = "pseudotime",
                     label_groups_by_cluster = FALSE,
                     label_leaves = FALSE)
pt <- monocle3::pseudotime(cds)
```

# Session Info

```{r}
sessionInfo()
```

